Periodograms for Multiband Astronomical Time Series

### 2015-08-18 writeup/paper.tex
more modifications for referee 2

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
\new{The underlying model of the Lomb-Scargle periodogram is nonlinear in frequency, and thus the likelihood surface in frequency is non-convex.
This non-convexity is readily apparent in the many local maxima of the periodogram, which makes it difficult to find the maximum via standard numerical optimization routines.
Thus in practice the global 
maximum of the periodogram is often found by a brute-force grid search \citep[for details see, e.g.][]{ICVG2014}.}
\old{The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and basis functions at different frequencies are not orthogonal. 
As a result, the periodogram has many local maxima. Thus in practice the global 
maximum of the periodogram is often found by a brute-force grid search (for details see, e.g. Ivezic 2014)}
A more general parametric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citet{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

@@

Second, notice that as the number of terms is increased, the general ``background'' level of the periodogram increases. This is due to the fact that the periodogram power is inversely related to the $\chi^2$ of the fit at each frequency. A more flexible higher-order model can better fit the data at all periods, not just the true period.
Thus in general the observed power of a higher-order Fourier model will be everywhere higher than the power of a lower-order Fourier model.
\new{One might hope that when adding terms, the correct-period model would show more of an improvement than the incorrect-period model (and thus the maximum would become more pronounced), but this does not generally hold.
Consider that in the extreme limit in which the number model parameters is equal to the number of data points, the model has enough flexibility to fit the data perfectly at {\it every} frequency, and the resulting periodogram would be a flat line!
Thus when adding more terms, the boost in power to the background must be larger on average than the boost in power to the signal.}

@@

First, we might consider ignoring band labels and simply compute a single standard Lomb-Scargle periodogram over the full dataset. This amounts to the assumption that a single sinusoid fits each band, and in practice will fail to account for astrophysical variability between bands.

Second, we might consider computing a standard Lomb-Scargle periodogram in each band, and using the additivity of $\chi^2$ along with \eq{chi2PN} to construct a multiband periodogram. This amounts to the assumption that the bands have completely independent phases and amplitudes, and has too many free parameters to be useful in most cases of interest (see \sect{relationship} for further discussion).}

To compute \old{such} a periodogram \new{which strikes a balance between these two extremes}, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The multiband model contains the following features:
\begin{enumerate}
  \item An $N_{base}$-term truncated Fourier fit which models \new{the shared variability among all $K$ observed bands}. \old{a latent parameter, which we'll call the ``overall variability''.}
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual within a single band from the \old{overall variability} \new{shared variability accounted for above}.
\end{enumerate}
The total number of parameters for $K$ bands is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  &y_k(t|\omega,\theta) =
  \theta_0 + \sum_{n=1}^{N_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right] +&\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{N_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.
\new{Note the potential for confusion: $N_{band}$ here is not the number of observed bands, but the number of Fourier components fit to the data each of the $K$ observed bands.}

@@

\subsection{Potential Improvements to the Multiband Method}
\new{A well-known (though often unrecognized) deficiency of Lomb-Scargle-type periodograms on unevenly-sampled data is that they do not measure the power of the signal in question, but the power of the signal {\it convolved with the observing with the survey window function}.
For regularly-sampled timeseries, this convolution is the source of the aliasing related to the Nyquist sampling limit; for non-regular sampling, this aliasing generally happens to some degree at {\it all} frequencies!
Thus even if the signal has a single well-defined period, this aliasing generally causes multiple maxima in the measured periodogram.
The multiband periodogram, as a generalization of Lomb-Scargle, shares this difficulty:}
\old{The fact that the multiband model does not select the top frequency each time points to its weaknesses: first, as a Fourier-like model,} it tends to respond to frequency structure in the window function as well as frequency structure in the data.
This can be viewed as a result of the very model simplicity which causes its success in the case of sparse multiband data: it cannot disentangle bias in the model from bias due to the survey window.

This could potentially be accounted for by correcting for the effect of the estimated window function; one potential method for this involves estimating the deconvolution of the window power and the observed power \citep{Roberts87}.
It may also be possible to propose a multiband extension of CARMA \citep{Kelly14}, or another forward-modeling approach to detecting periodicity.

>>>>

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
\old{The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and basis functions at different frequencies are not orthogonal. 
As a result, the periodogram has many local maxima.}
\new{The underlying model of the Lomb-Scargle periodogram is nonlinear in frequency, and thus the likelihood surface in frequency is non-convex.
This non-convexity is readily apparent in the many local maxima of the periodogram, which makes it difficult to find the maximum via standard numerical optimization routines.}
Thus in practice the global 
maximum of the periodogram is often found by a brute-force grid search \citep[for details see, e.g.][]{ICVG2014}.
A more general parametric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citet{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

@@

Second, notice that as the number of terms is increased, the general ``background'' level of the periodogram increases. This is due to the fact that the periodogram power is inversely related to the $\chi^2$ of the fit at each frequency. A more flexible higher-order model can better fit the data at all periods, not just the true period.
Thus in general the observed power of a higher-order Fourier model will be everywhere higher than the power of a lower-order Fourier model.
\new{One might hope that when adding terms, the correct-period model would show more of an improvement than the incorrect-period model (and thus the periodogram maximum would become more pronounced), but this does not generally hold.
Consider that in the extreme limit in which the number model parameters is equal to the number of data points, the model has enough flexibility to fit the data perfectly at {\it every} frequency, and the resulting periodogram would be a flat line at maximum power!
This can only be the case if, on average, addition of terms preferentially boosts the background level.}

@@

First, we might ignore band labels entirely and simply compute a single standard Lomb-Scargle periodogram over the full dataset. This amounts to the assumption that one global model suitably fits each band, and in practice will fail due to the astrophysical variability between bands.

Second, we might treat each band entirely independently and compute a standard Lomb-Scargle periodogram on each, and use the additivity of $\chi^2$ along with \eq{chi2PN} to construct a multiband periodogram. This amounts to the assumption that the bands have completely independent phases and amplitudes, and has too many free parameters to be useful in most cases of interest (see \sect{relationship} for further discussion).}

To compute \old{such} a periodogram \new{which strikes a balance between these two extremes}, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The multiband model presented here contains the following features:
\begin{enumerate}
  \item An $N_{base}$-term truncated Fourier fit which models \new{the shared variability among all $K$ observed bands}. \old{a latent parameter, which we'll call the ``overall variability''.}
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual within a single band from the \old{overall variability} \new{shared variability accounted for above}.
\end{enumerate}
The total number of parameters for $K$ bands is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  &y_k(t|\omega,\theta) =
  \theta_0 + \sum_{n=1}^{N_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right] +&\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{N_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.
\new{Note the potential for confusion: $N_{band}$ here is not the number of observed bands, but the number of Fourier components fit to the residuals in each of the $K$ observed bands.}

@@

\subsection{Potential Improvements to the Multiband Method}
\new{A well-known (though often unrecognized) deficiency of Lomb-Scargle-type periodograms on unevenly-sampled data is that they do not measure the power of the signal in question, but the power of the signal {\it convolved with the observing with the survey window function}.
For regularly-sampled timeseries, this convolution is the source of the aliasing related to the Nyquist sampling limit; for non-regular sampling, this aliasing generally happens to some degree at {\it all} frequencies!
Because of this, even a signal with a single well-defined period will result in a Lomb-Scargle periodogram with multiple maxima at locations which depend on both the underlying signal and the precise observing window.

The multiband periodogram, as a generalization of Lomb-Scargle, shares this difficulty:}
\old{The fact that the multiband model does not select the top frequency each time points to its weaknesses: first, as a Fourier-like model,} it tends to respond to frequency structure in the window function as well as frequency structure in the data.
This can be viewed as a result of the very model simplicity which causes its success in the case of sparse multiband data: it cannot disentangle bias in the model from bias due to the survey window.

This could potentially be accounted for by correcting for the effect of the estimated window function; one potential method for this involves estimating the deconvolution of the window power and the observed power \citep{Roberts87}.
It may also be possible to propose a multiband extension of, e.g., CARMA \citep{Kelly14} or another forward-modeling approach to detecting periodicity.
